---
title: "Computing REvoRate from Gradient Forest models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing REvoRate from Gradient Forest models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # set to TRUE if data files are available
)
```

## Overview

**REvoRate** (*Required Evolutionary Rate*) quantifies the minimum rate of
evolutionary change a population must achieve to remain adapted to its
shifting climate.  It is derived from the **genetic offset** computed by a
Gradient Forest (GF) model and a set of future climate raster stacks.

This vignette walks through the full workflow, mirroring the analysis in:

> Morando-Milà J, et al. (2025) Re-interpreting genetic offset: quantifying
> the least required evolutionary rate under climate change at the
> Mediterranean range margin of European beech. *Evolutionary Applications*.

---

## Installation

```{r install}
# From GitHub (recommended)
# install.packages("remotes")
remotes::install_github("your-github-username/REvoRateR")
```

---

## Prerequisites

You need:

1. A fitted `gradientForest` model object (here called `gfRef`).
2. A `RasterStack` of **present-day** climate variables, with layer names
   matching the GF predictor names (e.g. `bio2`, `bio4`, ...).
3. One `RasterStack` per **future period** (e.g. four 20-year windows
   2021–2040, 2041–2060, 2061–2080, 2081–2100).
4. Optionally, a `data.frame` of site coordinates (`Long`, `Lat`) with row
   names as site IDs.

```{r setup}
library(REvoRateR)
library(raster)
library(gradientForest)

# --- load your GF model and climate data ---
load("paperPrepGFmodel.RData")          # provides: gfRef, envGF, coord

# Present-day climate stack (layers named to match GF predictors)
current_stack <- stack(
  "wc2.1_30s_bio_2.tif",
  "wc2.1_30s_bio_4.tif",
  "wc2.1_30s_bio_5.tif",
  "wc2.1_30s_bio_8.tif",
  "wc2.1_30s_bio_9.tif",
  "wc2.1_30s_bio_18.tif"
)
names(current_stack) <- names(envGF)   # must match GF predictor names

# Future climate stacks — WorldClim CMIP6 (MPI-ESM1-2-HR, SSP5-8.5)
p1 <- stack("wc2.1_30s_bioc_MPI-ESM1-2-HR_ssp585_2021-2040.tif")
p2 <- stack("wc2.1_30s_bioc_MPI-ESM1-2-HR_ssp585_2041-2060.tif")
p3 <- stack("wc2.1_30s_bioc_MPI-ESM1-2-HR_ssp585_2061-2080.tif")
p4 <- stack("wc2.1_30s_bioc_MPI-ESM1-2-HR_ssp585_2081-2100.tif")

# Crop / mask future stacks to the same extent as current_stack
mask_r <- current_stack[[1]] > -100   # binary mask
for (s in c("p1","p2","p3","p4")) {
  tmp  <- get(s)
  names(tmp) <- paste0("BIO", seq_len(nlayers(tmp)))
  keep <- intersect(names(tmp), names(current_stack))
  tmp  <- tmp[[keep]]
  names(tmp) <- names(current_stack)[match(keep, names(current_stack))]
  if (!compareRaster(tmp, mask_r, stopiffalse = FALSE))
    tmp <- resample(tmp, mask_r, method = "bilinear")
  tmp <- crop(tmp, extent(mask_r))
  tmp <- mask(tmp, mask_r)
  assign(s, tmp)
}
```

---

## Running REvoRate()

```{r run}
result <- REvoRate(
  gf_model         = gfRef,
  current_stack    = current_stack,
  future_stacks    = list(p1 = p1, p2 = p2, p3 = p3, p4 = p4),
  period_midpoints = c(p1 = 2030, p2 = 2050, p3 = 2070, p4 = 2090),
  current_midpoint = 1985,       # midpoint of the 1970-2000 reference period
  coords           = coord,      # data.frame with Long, Lat; rownames = site IDs
  compute_climate_mismatch = TRUE,
  standardize_mismatch     = TRUE,
  verbose          = TRUE
)

print(result)
```

### What does `result` contain?

| Element | Description |
|---|---|
| `$genetic_offset` | Named list of `RasterLayer` — one per period |
| `$revoRate` | Named list of `RasterLayer` — one per period |
| `$climate_mismatch` | Named list of plain climate-mismatch rasters |
| `$climate_mismatch_gf` | Named list of GF-weighted mismatch rasters |
| `$delta_t` | Year-deltas used for each period |
| `$site_wide` | Wide-format data frame (one row per site) |
| `$site_long` | Long-format (tidy) data frame |

---

## Inspecting raster outputs

```{r rasters}
# Genetic offset for 2061-2080
plot(result$genetic_offset$p3, col = heat.colors(1000)[1000:1],
     main = "Genetic Offset 2061-2080")

# REvoRate for all periods
par(mfrow = c(2, 2))
for (nm in names(result$revoRate))
  plot(result$revoRate[[nm]], main = paste("REvoRate", nm))
```

---

## Tabular results

```{r tables}
head(result$site_wide)
head(result$site_long)

# Export
write.csv(result$site_wide, "REvoRate_wide.csv", row.names = FALSE)
write.csv(result$site_long, "REvoRate_long.csv", row.names = FALSE)
```

---

## Publication figure

```{r plot}
# Interactive preview
plot_revoRate(result, ncol = 6, palette = "cividis")

# Save as PDF
plot_revoRate(result,
              ncol      = 6,
              palette   = "cividis",
              save_path = "Fig_REvoRate_trajectories.pdf",
              width     = 12,
              height    = 9)
```

---

## Using individual functions

All steps are also available as standalone exported functions, so you can
integrate them into your own pipelines:

```{r standalone}
# Single-period genetic offset
go_p1 <- compute_genetic_offset(gfRef, current_stack, p1)

# REvoRate from a list of offsets
go_list <- list(p1 = go_p1,
                p2 = compute_genetic_offset(gfRef, current_stack, p2))
rr      <- compute_revoRate(go_list, delta_t = c(p1 = 45, p2 = 20))

# Climate mismatch
cm    <- compute_climate_mismatch_plain(current_stack, p1, standardize = TRUE)
cm_gf <- compute_climate_mismatch_gf(current_stack, p1, gfRef, standardize = TRUE)

# Extract at sites
extracted <- extract_site_metrics(
  coords           = coord,
  go_list          = go_list,
  rr_list          = rr,
  cm_plain         = list(p1 = cm),
  cm_gf            = list(p1 = cm_gf),
  period_midpoints = c(p1 = 2030, p2 = 2050),
  raster_crs       = raster::crs(rr$p1)
)
```

---

## Session info

```{r session}
sessionInfo()
```
